---
title: "Get started with DStressR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with DStressR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
```

`DStressR` models promoter-activity responses in high-throughput chemical
genomics screens. The package is designed as the stress-response counterpart to
[`DGrowthR`](https://bio-datascience.github.io/DGrowthR/): DGrowthR handles
growth-curve modeling, while DStressR handles promoter-compound effects after
accounting for growth and technical structure.

This vignette uses a small simulated screen so that the model-based named
workflow can be run without private data files. DStressR also exposes
compatibility workflows through the same entry point:
`workflow = "median_polish"` for the original median-polish p-value workflow
and `workflow = "empty_vector_control"` for the empty-vector-control workflow.

## Input files and table shape

DStressR expects a long measurement table with one row per measured
promoter-compound-replicate observation. In a complete rectangular screen, the
row count is approximately:

```text
n_promoters x (n_compound_wells + n_control_wells) x n_replicates
```

The table may contain additional rows when the same screen is repeated across
batches, library plates, measurement plates, or experimental days. The important
point is that these technical covariates remain in the table so they can be
modeled rather than accidentally averaged away.

The required biological and technical information is:

- promoter or reporter construct
- compound or library-well identifier
- negative-control compound, usually `DMSO`
- luminescence summary
- growth summary
- replicate and optional technical covariates such as batch or plate

Column names are not fixed. They are mapped explicitly in `prepare_assay()`.

For the original Campylobacter promoter-library workflow, the convenience helper
`read_campylobacter_expression()` joins two exported files:

- `expression_values.tsv.gz`: one row per promoter-library-well-replicate
  observation. It must contain either `srn_code` or both `libplate` and `well`,
  plus the promoter, luminescence, growth, and technical columns used
  downstream.
- `LibMap.txt`: one row per compound/library well with columns `Library plate`,
  `Well`, `ProductName`, and `Catalog Number`.

The helper reconstructs the library-well key as
`paste0("lp", Library plate, "_", Well)` for `LibMap.txt`, joins the compound
annotations onto the expression table, and returns a long table with the same
number of rows as `expression_values.tsv.gz`.

```{r campylobacter-import, eval = FALSE}
expression_df <- read_campylobacter_expression(
  expression_file = "expression_values.tsv.gz",
  libmap_file = "LibMap.txt"
)

assay <- prepare_assay(
  expression_df,
  promoter = "promoter",
  compound = "srn_code",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  batch = "batch",
  plate = "libplate",
  replicate = "replicate"
)
```

## Simulate a chemical-genomics screen

```{r setup}
library(DStressR)

screen <- simulate_screen(
  n_promoters = 12,
  n_compounds = 40,
  n_replicates = 3,
  seed = 1
)

head(screen)
```

The simulated table contains one row per promoter-compound-replicate well. It
includes luminescence, growth, batch, replicate, and hidden truth columns used
only to check the example.

```{r columns}
names(screen)
```

## Prepare the assay

`prepare_assay()` converts luminescence and growth summaries into a
growth-adjusted log2 response. By default, DStressR estimates promoter-specific
growth exponents from DMSO control wells and shrinks them toward a global
control-well slope.

```{r prepare}
assay <- prepare_assay(
  screen,
  promoter = "promoter",
  compound = "compound",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  batch = "batch",
  replicate = "replicate"
)

attr(assay, "destress")$growth_exponent_fit
```

To reproduce the older fixed-ratio style response,
`growth_exponent = 1` gives the familiar `log2(luminescence / growth)` scale.

```{r fixed-alpha}
assay_fixed <- prepare_assay(
  screen,
  promoter = "promoter",
  compound = "compound",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  growth_exponent = 1,
  batch = "batch",
  replicate = "replicate"
)
```

## Fit the model workflow

`fit_workflow(..., workflow = "model")` fits promoter and compound effects
while accounting for technical covariates. The result table reports both the
DMSO-relative total effect and the promoter-specific effect after subtracting
the compound-wide effect.

```{r fit}
fit <- fit_workflow(
  assay,
  workflow = "model",
  technical = c("batch", "replicate"),
  empirical_bayes = TRUE
)

tab <- results(fit)
tab <- adjust_pvalues(tab)

head(tab)
```

The key columns are:

- `total_effect`: DMSO-relative response for the promoter-compound pair.
- `global_effect`: average compound-wide response across promoters.
- `specific_effect`: promoter-specific deviation from the compound-wide effect.
- `specific_pvalue` and `specific_padj`: test and BH-adjusted p-value.

The direct fitting functions remain available for existing scripts, but
`fit_workflow()` makes the selected statistical path explicit and is the
recommended entry point for new analyses.

## Call hits

```{r hits}
hits <- call_hits(
  tab,
  fdr = 0.05,
  lfc = 0.5,
  effect = "specific_effect",
  padj = "specific_padj"
)

table(hits$hit)

head(
  hits[order(hits$specific_padj, -abs(hits$specific_effect)), ],
  10
)
```

In real screens, the FDR cutoff should be paired with domain knowledge and
diagnostics. For DMSO-rich designs, empirical replicate/permutation p-values can
be used as an additional calibration check.

## Visualize the screen

The standard volcano plot labels the strongest promoter-compound pairs.

```{r volcano, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_volcano(
  hits,
  effect = "specific_effect",
  padj = "specific_padj",
  fdr = 0.05,
  lfc = 0.5,
  top_n = 8,
  top_promoters = 6
)
```

A response heatmap gives a matrix-level view of the promoter-compound response
surface.

```{r heatmap, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_response_heatmap(
  hits,
  value = "specific_effect",
  top_n_compounds = 30
)
```

For broader screens, clustered heatmaps help reveal compound and promoter
groups with similar response profiles.

```{r clustered, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_response_clustered_heatmap(
  hits,
  value = "specific_effect",
  top_n_compounds = 30,
  title = "Clustered DStressR response map"
)
```

## Optional DGrowthR handoff

The current default hit model uses the exported growth summary column supplied
to `prepare_assay()`. If growth curves have already been modeled with DGrowthR,
the DGrowthR-derived growth parameter can be joined explicitly before assay
preparation.

```{r dgrowthr-handoff, eval = FALSE}
screen2 <- add_dgrowthr_growth(
  screen,
  object = dgrowthr_fit,
  by = "curve_id",
  model_covariate = "curve_id",
  growth_metric = "OD_16",
  output = "dgrowthr_od16"
)

assay2 <- prepare_assay(
  screen2,
  promoter = "promoter",
  compound = "compound",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "dgrowthr_od16",
  batch = "batch",
  replicate = "replicate"
)
```

This explicit handoff keeps the selected DStressR workflow reproducible, while
making DGrowthR-based sensitivity analyses straightforward.
