---
title: "Get started with DStressR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with DStressR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
```

`DStressR` models promoter-activity responses in high-throughput chemical
genomics screens. The package is designed as the stress-response counterpart to
[`DGrowthR`](https://bio-datascience.github.io/DGrowthR/): DGrowthR handles
growth-curve modeling, while DStressR handles promoter-compound effects after
accounting for growth and technical structure.

This vignette uses a small simulated screen so that the full workflow can be run
without private data files.

## Simulate a chemical-genomics screen

```{r setup}
library(DStressR)

screen <- simulate_screen(
  n_promoters = 12,
  n_compounds = 40,
  n_replicates = 3,
  seed = 1
)

head(screen)
```

The simulated table contains one row per promoter-compound-replicate well. It
includes luminescence, growth, batch, replicate, and hidden truth columns used
only to check the example.

```{r columns}
names(screen)
```

## Prepare the assay

`prepare_assay()` converts luminescence and growth summaries into a
growth-adjusted log2 response. By default, DStressR estimates promoter-specific
growth exponents from DMSO control wells and shrinks them toward a global
control-well slope.

```{r prepare}
assay <- prepare_assay(
  screen,
  promoter = "promoter",
  compound = "compound",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  batch = "batch",
  replicate = "replicate"
)

attr(assay, "destress")$growth_exponent_fit
```

To reproduce the older fixed-ratio style response,
`growth_exponent = 1` gives the familiar `log2(luminescence / growth)` scale.

```{r fixed-alpha}
assay_fixed <- prepare_assay(
  screen,
  promoter = "promoter",
  compound = "compound",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  growth_exponent = 1,
  batch = "batch",
  replicate = "replicate"
)
```

## Fit the DStressR model

`fit_destress()` fits promoter and compound effects while accounting for
technical covariates. The result table reports both the DMSO-relative total
effect and the promoter-specific effect after subtracting the compound-wide
effect.

```{r fit}
fit <- fit_destress(
  assay,
  technical = c("batch", "replicate"),
  empirical_bayes = TRUE
)

tab <- results(fit)
tab <- adjust_pvalues(tab)

head(tab)
```

The key columns are:

- `total_effect`: DMSO-relative response for the promoter-compound pair.
- `global_effect`: average compound-wide response across promoters.
- `specific_effect`: promoter-specific deviation from the compound-wide effect.
- `specific_pvalue` and `specific_padj`: test and BH-adjusted p-value.

## Call hits

```{r hits}
hits <- call_hits(
  tab,
  fdr = 0.05,
  lfc = 0.5,
  effect = "specific_effect",
  padj = "specific_padj"
)

table(hits$hit)

head(
  hits[order(hits$specific_padj, -abs(hits$specific_effect)), ],
  10
)
```

In real screens, the FDR cutoff should be paired with domain knowledge and
diagnostics. For DMSO-rich designs, empirical replicate/permutation p-values can
be used as an additional calibration check.

## Visualize the screen

The standard volcano plot labels the strongest promoter-compound pairs.

```{r volcano, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_volcano(
  hits,
  effect = "specific_effect",
  padj = "specific_padj",
  fdr = 0.05,
  lfc = 0.5,
  top_n = 8,
  top_promoters = 6
)
```

A response heatmap gives a matrix-level view of the promoter-compound response
surface.

```{r heatmap, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_response_heatmap(
  hits,
  value = "specific_effect",
  top_n_compounds = 30
)
```

For broader screens, clustered heatmaps help reveal compound and promoter
groups with similar response profiles.

```{r clustered, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_response_clustered_heatmap(
  hits,
  value = "specific_effect",
  top_n_compounds = 30,
  title = "Clustered DStressR response map"
)
```

## Optional DGrowthR handoff

The current default hit model uses the exported growth summary column supplied
to `prepare_assay()`. If growth curves have already been modeled with DGrowthR,
the DGrowthR-derived growth parameter can be joined explicitly before assay
preparation.

```{r dgrowthr-handoff, eval = FALSE}
screen2 <- add_dgrowthr_growth(
  screen,
  object = dgrowthr_fit,
  by = "curve_id",
  model_covariate = "curve_id",
  growth_metric = "OD_16",
  output = "dgrowthr_od16"
)

assay2 <- prepare_assay(
  screen2,
  promoter = "promoter",
  compound = "compound",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "dgrowthr_od16",
  batch = "batch",
  replicate = "replicate"
)
```

This explicit handoff keeps the original DStressR hit workflow reproducible,
while making DGrowthR-based sensitivity analyses straightforward.
