---
title: "Get started with DStressR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with DStressR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
```

`DStressR` models reporter-activity responses in high-throughput chemical
genomics screens. The package is designed as the stress-response counterpart to
[`DGrowthR`](https://bio-datascience.github.io/DGrowthR/): DGrowthR handles
growth-curve modeling, while DStressR handles reporter-perturbation effects after
accounting for growth and technical structure.

This vignette uses a small simulated screen so that the model-based named
workflow can be run without private data files. DStressR also exposes
compatibility workflows through the same entry point:
`workflow = "median_polish"` for the original median-polish p-value workflow
and `workflow = "empty_vector_control"` for the empty-vector-control workflow.

## Input files and table shape

DStressR expects a long measurement table with one row per measured
reporter-perturbation-replicate observation. In a complete rectangular screen, the
row count is approximately:

```text
n_reporters x (n_compound_wells + n_control_wells) x n_replicates
```

The table may contain additional rows when the same screen is repeated across
batches, library plates, measurement plates, or experimental days. The important
point is that these technical covariates remain in the table so they can be
modeled rather than accidentally averaged away.

The required biological and technical information is:

- reporter or reporter construct
- perturbation or library-well identifier
- negative-control perturbation, usually `DMSO`
- luminescence summary
- growth summary
- replicate and optional technical covariates such as batch or plate

Column names are not fixed. They are mapped explicitly in `prepare_assay()`.

For the original Campylobacter reporter-library workflow, the convenience helper
`read_campylobacter_expression()` joins two exported files:

- `expression_values.tsv.gz`: one row per reporter-library-well-replicate
  observation. It must contain either `srn_code` or both `libplate` and `well`,
  plus the reporter, luminescence, growth, and technical columns used
  downstream.
- `LibMap.txt`: one row per perturbation/library well with columns `Library plate`,
  `Well`, `ProductName`, and `Catalog Number`.

The helper reconstructs the library-well key as
`paste0("lp", Library plate, "_", Well)` for `LibMap.txt`, joins the perturbation
annotations onto the expression table, and returns a long table with the same
number of rows as `expression_values.tsv.gz`.

```{r campylobacter-import, eval = FALSE}
expression_df <- read_campylobacter_expression(
  expression_file = "expression_values.tsv.gz",
  libmap_file = "LibMap.txt"
)

assay <- prepare_assay(
  expression_df,
  reporter = "reporter",
  perturbation = "srn_code",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  batch = "batch",
  plate = "libplate",
  replicate = "replicate"
)
```

## Simulate a chemical-genomics screen

```{r setup}
library(DStressR)

screen <- simulate_screen(
  n_reporters = 12,
  n_perturbations = 40,
  n_replicates = 3,
  seed = 1
)

head(screen)
```

The simulated table contains one row per reporter-perturbation-replicate well. It
includes luminescence, growth, batch, replicate, and hidden truth columns used
only to check the example.

```{r columns}
names(screen)
```

## Prepare the assay

`prepare_assay()` converts luminescence and growth summaries into a
growth-adjusted log2 response. By default, DStressR estimates reporter-specific
growth exponents from DMSO control wells and shrinks them toward a global
control-well slope.

```{r prepare}
assay <- prepare_assay(
  screen,
  reporter = "reporter",
  perturbation = "perturbation",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  batch = "batch",
  replicate = "replicate"
)

attr(assay, "destress")$growth_exponent_fit
```

To reproduce the older fixed-ratio style response,
`growth_exponent = 1` gives the familiar `log2(luminescence / growth)` scale.

```{r fixed-alpha}
assay_fixed <- prepare_assay(
  screen,
  reporter = "reporter",
  perturbation = "perturbation",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  growth_exponent = 1,
  batch = "batch",
  replicate = "replicate"
)
```

## Fit the model workflow

`fit_destress()` fits reporter and perturbation effects while accounting for
technical covariates. The result table reports both the DMSO-relative total
effect and the reporter-specific effect after subtracting the perturbation-wide
effect.

```{r fit}
fit <- fit_destress(
  assay,
  technical = c("batch", "replicate"),
  empirical_bayes = TRUE
)

tab <- results(fit)
tab <- adjust_pvalues(tab)

head(tab)
```

The key columns are:

- `total_effect`: DMSO-relative response for the reporter-perturbation pair.
- `global_effect`: average perturbation-wide response across reporters.
- `specific_effect`: reporter-specific deviation from the perturbation-wide effect.
- `specific_pvalue` and `specific_padj`: test and BH-adjusted p-value.

The direct fitting functions remain available for existing scripts, but
`fit_destress()` is the recommended entry point for new model-based analyses.

## Optional background reporter calibration

If a screen contains a matched background reporter, such as an Empty Vector
Control, the background reporter can be used during response construction. When
`background_reporter` is supplied, DStressR uses Huber calibration by default;
least-squares calibration and direct subtraction remain available through
`background_method`.

```{r background-calibration, eval = FALSE}
assay_bg <- prepare_assay(
  screen_with_evc,
  reporter = "reporter",
  perturbation = "perturbation",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  batch = "batch",
  replicate = "replicate",
  background_reporter = "EVC",
  background_by = c("perturbation", "batch", "replicate")
)

fit_bg <- fit_destress(
  assay_bg,
  technical = c("batch", "replicate"),
  empirical_bayes = TRUE
)
```

## Call hits

```{r hits}
hits <- call_hits(
  tab,
  fdr = 0.05,
  lfc = 0.5,
  effect = "specific_effect",
  padj = "specific_padj"
)

table(hits$hit)

head(
  hits[order(hits$specific_padj, -abs(hits$specific_effect)), ],
  10
)
```

In real screens, the FDR cutoff should be paired with domain knowledge and
diagnostics. For DMSO-rich designs, empirical replicate/permutation p-values can
be used as an additional calibration check.

## Visualize the screen

The standard volcano plot labels the strongest reporter-perturbation pairs.

```{r volcano, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_volcano(
  hits,
  effect = "specific_effect",
  padj = "specific_padj",
  fdr = 0.05,
  lfc = 0.5,
  top_n = 8,
  top_reporters = 6
)
```

A response heatmap gives a matrix-level view of the reporter-perturbation response
surface.

```{r heatmap, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_response_heatmap(
  hits,
  value = "specific_effect",
  top_n_perturbations = 30
)
```

For broader screens, clustered heatmaps help reveal perturbation and reporter
groups with similar response profiles.

```{r clustered, eval = requireNamespace("ggplot2", quietly = TRUE)}
plot_response_clustered_heatmap(
  hits,
  value = "specific_effect",
  top_n_perturbations = 30,
  title = "Clustered DStressR response map"
)
```

## Optional DGrowthR handoff

The current default hit model uses the exported growth summary column supplied
to `prepare_assay()`. If growth curves have already been modeled with DGrowthR,
the DGrowthR-derived growth parameter can be joined explicitly before assay
preparation.

```{r dgrowthr-handoff, eval = FALSE}
screen2 <- add_dgrowthr_growth(
  screen,
  object = dgrowthr_fit,
  by = "curve_id",
  model_covariate = "curve_id",
  growth_metric = "OD_16",
  output = "dgrowthr_od16"
)

assay2 <- prepare_assay(
  screen2,
  reporter = "reporter",
  perturbation = "perturbation",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "dgrowthr_od16",
  batch = "batch",
  replicate = "replicate"
)
```

This explicit handoff keeps the selected DStressR workflow reproducible, while
making DGrowthR-based sensitivity analyses straightforward.
