---
title: "E. coli fluorescence stress screen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{E. coli fluorescence stress screen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
```

This article documents how DStressR can be used for a fluorescence/OD
reporter-stress experiment. The example is based on the Dryad data associated
with Dash et al. (2024), *mSystems*, which contain time-resolved GFP and OD
measurements for *E. coli* global-regulator reporters under weak-stress
conditions.

The Dash/Dryad raw data are not shipped with the package. The complete local
workflow is kept in the repository under `analysis/dryad_global_regulators/`.
After downloading and unzipping the Dryad archive, the expected local layout is:

```text
analysis/dryad_global_regulators/raw/Data/
```

The main script for the windowed weak-stress analysis is:

```sh
Rscript analysis/dryad_global_regulators/run_weak_stress_window_calibrated_pipeline.R
```

## Analysis Table

The weak-stress fluorescence time series are summarized into GFP AUC values and
matched OD AUC values. The analysis table should use source-appropriate column
names. In this dataset, the DStressR perturbation variable is a `stress`, not a
drug or compound.

```{r dryad-table, eval = FALSE}
weak_stress <- read.delim(
  "analysis/outputs/dryad_global_regulators/dryad_weak_stress_windows_calibrated_alpha_input.tsv",
  check.names = FALSE
)

weak_stress$stress <- weak_stress$compound
weak_stress$compound <- NULL

names(weak_stress)
```

The reporter unit can be a biological reporter or a derived reporter-window
unit. For the analysis used during manuscript development, each base reporter is
crossed with a time window:

```{r dryad-reporter-window, eval = FALSE}
weak_stress$pseudo_reporter <- paste(
  weak_stress$promoter,
  weak_stress$window,
  sep = " | "
)
```

## DStressR Model

The response model maps the Dash data to the generic DStressR interface by
using `reporter = "pseudo_reporter"` and `perturbation = "stress"`. Standard
growth is the reference condition. The fluorescence signal is represented by
`gfp_auc`, and growth by `od_auc`.

```{r dryad-destress, eval = FALSE}
alpha <- read.delim(
  "analysis/outputs/dryad_global_regulators/dryad_weak_stress_windows_calibrated_alpha_supplied.tsv",
  check.names = FALSE
)
stress_alpha <- setNames(alpha$alpha_calibrated, alpha$pseudo_reporter)

assay <- prepare_assay(
  weak_stress,
  reporter = "pseudo_reporter",
  perturbation = "stress",
  control = "Standard",
  lux = "gfp_auc",
  growth = "od_auc",
  growth_exponent = stress_alpha,
  replicate = "replicate"
)

fit <- fit_destress(
  assay,
  technical = "replicate",
  empirical_bayes = TRUE,
  adjustment = "by_reporter",
  interaction = FALSE
)

tab <- results(fit)
hits <- call_hits(
  tab,
  fdr = 0.05,
  effect = "specific_effect",
  padj = "specific_padj_by_reporter"
)
```

The same workflow can be run with `growth_exponent = 1` when the goal is to
match a GFP/OD-style response more directly. The calibrated-alpha analysis shows
how DStressR can use additional no-stress growth-transition data to estimate the
response scale, while the weak-stress subset is used for inference.

## Reproducibility Scripts

The repository scripts write intermediate inputs, fitted results, p-value
histograms, volcano plots, and effect heatmaps under:

```text
analysis/outputs/dryad_global_regulators/
```

The most relevant scripts are:

```text
analysis/dryad_global_regulators/run_weak_stress_window_pipeline.R
analysis/dryad_global_regulators/run_weak_stress_window_calibrated_pipeline.R
analysis/dryad_global_regulators/run_weak_stress_window_rank1_adjustment.R
analysis/dryad_global_regulators/plot_weak_stress_three_target_inference.R
```

This example is intentionally different from the Binsfeld reporter-drug screen:
it illustrates a small structured fluorescence/OD reporter-stress design rather
than a large chemical library.
