---
title: "E. coli reporter-perturbation screen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{E. coli reporter-perturbation screen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
```

This article describes the public *E. coli* reporter screen used as the package
application for DStressR. The data are from Binsfeld et al. (2025), PLOS
Biology, and are distributed with the package as compact, documented R data
objects. The AUC-level screen is available as `binsfeld_reporter_auc`; the
matching author score and Z-score table is available as
`binsfeld_reporter_scores`.

The source files are the PLOS Biology supplementary data and the associated
Zenodo archive:

- article: <https://doi.org/10.1371/journal.pbio.3003260>
- archive: <https://doi.org/10.5281/zenodo.15600688>

## Public data objects

```{r data-overview}
binsfeld_reporter_auc <- DStressR::binsfeld_reporter_auc
binsfeld_reporter_scores <- DStressR::binsfeld_reporter_scores

dim(binsfeld_reporter_auc)
table(binsfeld_reporter_auc$strain)
length(unique(binsfeld_reporter_auc$reporter))
length(unique(binsfeld_reporter_auc$drug))
```

The AUC table is a long table with one row per strain, reporter, drug or
control, replicate, and serial-dilution observation. The `drug` column keeps the
original Binsfeld labels, including `Water_1` and `Water_2`. The derived
`compound` column is retained only as a convenience grouping variable for
summaries.
The source `concentration_index` increases with dilution; `dose_level` reverses
this coding so that larger values correspond to higher concentration for
non-water compound wells. For water controls, the same column records the
matched concentration-index position in the screen layout.

```{r auc-columns}
names(binsfeld_reporter_auc)
```

The score table stores the public author scores and Z-scores in long form. It
is used by the reproducibility workflow to reconstruct the reference hit rule.

```{r score-columns}
names(binsfeld_reporter_scores)
```

## DStressR analysis

For the package application, the wild-type reporter data are filtered to rows
that passed the original quality-control flag. The original `Water_1` and
`Water_2` labels are supplied together as the reference condition. The Empty
Vector Control reporter is supplied to `prepare_assay()` as a matched
background reporter, so the default response model performs Huber-calibrated
background adjustment before model-based testing.

```{r ecoli-destress, eval = FALSE}
wt_auc <- subset(
  binsfeld_reporter_auc,
  strain == "WT" & removed == "No"
)

assay <- prepare_assay(
  wt_auc,
  reporter = "reporter",
  perturbation = "drug",
  control = c("Water_1", "Water_2"),
  lux = "lux_auc",
  growth = "od_auc",
  growth_exponent = "estimate",
  batch = "dose_level",
  replicate = "replicate",
  growth_covariates = "replicate",
  numeric_covariates = "dose_level",
  background_reporter = "EVC",
  background_by = c("drug", "dose_level", "replicate")
)

fit <- fit_destress(
  assay,
  preset = "model",
  technical = c("replicate", "dose_level"),
  empirical_bayes = TRUE,
  adjustment = "by_reporter",
  interaction = FALSE
)

hits <- call_hits(
  results(fit),
  fdr = 0.05,
  effect = "specific_effect"
)
```

The full comparison workflow is kept outside the package vignette build because
it creates manuscript-scale figures and tables. From the repository root, run:

```sh
Rscript analysis/ecoli_promoter_screen/run_evc_calibrated_analysis.R
```

The workflow compares three hit sets:

- the reconstructed Binsfeld-style Wilcoxon/Z-score reference rule
- DStressR without EV
- DStressR with EV

The current wild-type comparison gives:

```text
Reference WT hits: 53
DStressR without EV WT hits: 80
DStressR with EV WT hits: 92
All three analyses: 35
Reference-only hits: 16
DStressR-without-EV only hits: 9
DStressR-with-EV only hits: 19
Both DStressR workflows only: 36
Union significant by at least one primary analysis: 117
```

The full interaction-model sensitivity analysis used in the manuscript
appendix can be reproduced with:

```sh
Rscript analysis/ecoli_promoter_screen/run_interaction_sensitivity.R
```

In the current wild-type analysis, the interaction model calls 56 hits with EV
calibration and 51 hits without EV calibration.

Additional scripts in `analysis/ecoli_promoter_screen/` recreate the
two-method report, volcano plots, heatmaps, Venn diagrams, and response-modeling
figures used during manuscript development. Generated outputs are written under
`analysis/outputs/` and are intentionally ignored by Git.
