---
title: "Original median-polish workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Original median-polish workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
library(DStressR)
```

`DStressR` includes a legacy-compatible median-polish path so that the original
workflow can be reproduced before moving to the model-based DStressR analyses.
This is useful for regression testing, method comparison, and explaining which
parts of the new framework are replacements rather than silent changes.

The function that implements this baseline is `fit_median_polish()`.

## Required input files

For the original Campylobacter promoter-library workflow, the median-polish
baseline starts from two exported files.

### `expression_values.tsv.gz`

This is a long table with one row per measured
promoter-library-well-replicate observation. Its shape is approximately:

```text
n_promoters x (n_compound_wells + n_dmso_wells + n_noisy_dmso_wells) x n_replicates
```

The required columns are:

- promoter identifier, for example `promoter`
- compound/library-well identifier, for example `srn_code`
- library plate, for example `libplate`
- replicate, for example `replicate`
- growth-normalized log2 luminescence response, originally
  `log2.auc.16hmeasured.normed`

If `srn_code` is not already present, `read_campylobacter_expression()` can
reconstruct it from `libplate` and `well`.

### `LibMap.txt`

This is a library annotation table with one row per compound/library well. The
required columns are:

- `Library plate`
- `Well`
- `ProductName`
- `Catalog Number`

The original workflow identifies ordinary DMSO controls and noisy DMSO wells
from this table. DStressR uses the same compound key:

```r
srn_code = paste0("lp", `Library plate`, "_", Well)
```

## Reproduce the original workflow

```{r original-files, eval = FALSE}
library(DStressR)

expression_df <- read_campylobacter_expression(
  expression_file = "expression_values.tsv.gz",
  libmap_file = "LibMap.txt"
)

libmap <- read.delim("LibMap.txt", check.names = FALSE)
libmap$srn_code <- paste0("lp", libmap[["Library plate"]], "_", libmap[["Well"]])

dmso_srn_codes <- libmap$srn_code[libmap$ProductName == "DMSO"]
dmso_noisy_srn_codes <- libmap$srn_code[libmap$ProductName == "DMSO noisy"]

legacy <- fit_median_polish(
  expression_df,
  promoter = "promoter",
  compound = "srn_code",
  libplate = "libplate",
  replicate = "replicate",
  response = "log2.auc.16hmeasured.normed",
  control = dmso_srn_codes,
  exclude = dmso_noisy_srn_codes,
  maxiter = 1000,
  eps = 1e-8
)
```

The replicate-level table corresponds to the original
`expression_df.pvalues.tsv.gz` output:

```{r replicate-output, eval = FALSE}
replicate_pvalues <- legacy$replicate_results

head(replicate_pvalues)
```

The pair-level table corresponds to the original hit-calling table: DMSO and
excluded controls are removed, the largest replicate-level p-value is retained
for each promoter-compound pair, and BH correction is applied within promoter.

```{r pair-output, eval = FALSE}
hit_table <- legacy$pair_results

head(hit_table)
table(hit_table$hit)
```

## What the function does

`fit_median_polish()` follows the original hit-determination steps:

1. Build `promoter_libplate_replicate = paste(promoter, libplate, replicate, sep = "_")`.
2. Compute the DMSO mean normalized response within each
   promoter-libplate-replicate group.
3. Compute `log2FC = normalized.lux - avg.dmso.expression`.
4. Form a matrix with promoter-libplate-replicate groups in rows and
   `srn_code` compound/library wells in columns.
5. Apply `stats::medpolish(..., na.rm = TRUE, maxiter = 1000, eps = 1e-8)`.
6. Use the median-polish residuals as `log2FC.polished`.
7. Estimate the DMSO residual mean and standard deviation within each
   promoter-libplate-replicate group.
8. Compute
   `zscore = (log2FC.polished - dmso.avg_dmsoFC) / dmso.stdv_dmsoFC`.
9. Compute two-sided Gaussian p-values.
10. For hit calling, remove DMSO/noisy-DMSO wells, keep the largest p-value
    across replicates for each promoter-compound pair, and apply BH correction
    within promoter.

## Minimal executable example

This toy example is deliberately tiny; it only shows the mechanics of the
legacy path. Real screens need multiple DMSO wells per
promoter-libplate-replicate group to estimate the DMSO null distribution.

```{r toy}
toy <- expand.grid(
  promoter = c("P1", "P2"),
  libplate = "lp1",
  replicate = c("r1", "r2"),
  srn_code = c("DMSO1", "DMSO2", "C1", "C2"),
  stringsAsFactors = FALSE
)

toy$log2.auc.16hmeasured.normed <- c(
  10.0, 10.2, 11.4, 9.4,
  10.1, 10.3, 11.5, 9.5,
  12.0, 12.2, 13.6, 11.6,
  12.1, 12.3, 13.7, 11.7
)

legacy <- fit_median_polish(
  toy,
  control = c("DMSO1", "DMSO2")
)

legacy$replicate_results
legacy$pair_results
```

## Compare against DStressR model-based inference

The median-polish path is intended as a reproducible baseline. The model-based
DStressR analysis starts from the same long table but uses `prepare_assay()`
and `fit_destress()`:

```{r model-path, eval = FALSE}
assay <- prepare_assay(
  expression_df,
  promoter = "promoter",
  compound = "srn_code",
  control = "DMSO",
  lux = "LUX.AUC_16",
  growth = "od_16h.measured",
  batch = "batch",
  plate = "libplate",
  replicate = "replicate"
)

fit <- fit_destress(
  assay,
  technical = c("batch", "libplate", "replicate"),
  empirical_bayes = TRUE
)

model_results <- results(fit)
```

This gives a direct apples-to-apples setup: `fit_median_polish()` recovers the
legacy workflow, while `fit_destress()` provides the model-based replacement.
