---
title: "Salmonella Empty Vector workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Salmonella Empty Vector workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8
)
library(DStressR)
```

The Salmonella StressRegNet workflow uses Empty Vector Control (EVC) reporter
strains to estimate compound-specific background Lux signal. In the original
analysis, the `PEVC3` reporter is used as the reference. For each compound,
the two PEVC3 replicate values are averaged, and this compound-specific EVC
average is subtracted from every reporter strain.

This is different from the Campylobacter median-polish baseline. The Salmonella
baseline is:

```text
growth-normalized Lux
  -> promoter-replicate median centering
  -> subtract compound-specific PEVC3 average
  -> estimate DMSO null per promoter-replicate
  -> z-test p-values
  -> retain largest replicate p-value per promoter-compound
  -> BH adjustment within promoter
```

DStressR implements this compatibility path with `fit_empty_vector_control()`.

## Input files

The workflow starts from the processed Salmonella Lux-estimation output:

### `lux_auc_filtered_median.tsv.gz`

This is a long table with one row per promoter-library-well-replicate
observation. Its rows are approximately:

```text
n_promoters x n_library_wells x n_replicates
```

after growth filtering and reporter-quality filtering. The required columns are:

- `promoter`
- `srn_code`
- `replicate`
- `log2.lux.normed.centered`

The response column `log2.lux.normed.centered` is already:

1. Lux AUC until 10 hours,
2. divided by the OD reached at 10 hours,
3. log2-transformed,
4. median-centered within promoter-replicate.

### `LibMap.tsv.gz`

This file annotates library wells and identifies DMSO controls. Required columns
are:

- `Library plate`
- `New well`
- `Catalog Number`
- `ProductName`

The compound key used by the workflow is:

```r
srn_code = paste(sub("LibPlate", "lp", `Library plate`), `New well`, sep = "_")
```

## Reproduce the original Salmonella workflow

```{r original-files, eval = FALSE}
expression_df <- read.delim(gzfile("lux_auc_filtered_median.tsv.gz"),
                            check.names = FALSE)
libmap <- read.delim(gzfile("LibMap.tsv.gz"), check.names = FALSE)

libmap$libplate <- sub("LibPlate", "lp", libmap[["Library plate"]])
libmap$srn_code <- paste(libmap$libplate, libmap[["New well"]], sep = "_")

dmso_srn_codes <- libmap$srn_code[libmap[["Catalog Number"]] == "DMSO"]
dmso_noisy_srn_codes <- libmap$srn_code[libmap[["Catalog Number"]] == "DMSO noisy"]

evc <- fit_empty_vector_control(
  expression_df,
  promoter = "promoter",
  compound = "srn_code",
  replicate = "replicate",
  response = "log2.lux.normed.centered",
  empty_vector_promoter = "PEVC3",
  control = dmso_srn_codes,
  exclude = dmso_noisy_srn_codes,
  remove_promoters = "PmgrR"
)
```

The replicate-level table contains the EVC-normalized fold change, DMSO null
parameters, z-scores, and raw p-values:

```{r replicate-output, eval = FALSE}
replicate_pvalues <- evc$replicate_results

head(replicate_pvalues)
```

The pair-level table reproduces the original conservative aggregation: for each
promoter-compound pair, retain the largest replicate-level p-value and apply
BH adjustment within promoter.

```{r pair-output, eval = FALSE}
hit_table <- evc$pair_results

head(hit_table)
table(hit_table$hit)
```

## What the function does

`fit_empty_vector_control()` follows the original Salmonella hit-determination
logic:

1. Remove noisy DMSO wells and failed reporters, for example `PmgrR`.
2. Keep the already centered response column, usually
   `log2.lux.normed.centered`.
3. For each `srn_code`, compute the average PEVC3 response across PEVC3
   replicates.
4. Compute
   `log.evcfc = expression_value - average_PEVC3_expression_for_srn_code`.
5. Remove the PEVC3 reference rows from the tested reporter table.
6. Gather DMSO `log.evcfc` values within each promoter-replicate group.
7. Estimate `dmso.mean` and `dmso.stdv`.
8. Compute
   `zscore = (log.evcfc - dmso.mean) / dmso.stdv`.
9. Compute two-sided Gaussian p-values.
10. Retain the largest replicate-level p-value for each promoter-compound pair.
11. Apply BH correction within promoter and label hits.

## Minimal executable example

```{r toy}
toy <- expand.grid(
  promoter = c("PEVC3", "P1", "P2"),
  replicate = c("r1", "r2"),
  srn_code = c("DMSO1", "DMSO2", "C1"),
  stringsAsFactors = FALSE
)

toy$value <- NA_real_
toy$value[toy$promoter == "PEVC3" & toy$srn_code == "DMSO1"] <- c(1.0, 1.2)
toy$value[toy$promoter == "PEVC3" & toy$srn_code == "DMSO2"] <- c(1.1, 1.3)
toy$value[toy$promoter == "PEVC3" & toy$srn_code == "C1"] <- c(2.0, 2.2)

toy$value[toy$promoter == "P1" & toy$srn_code == "DMSO1"] <- c(1.5, 1.7)
toy$value[toy$promoter == "P1" & toy$srn_code == "DMSO2"] <- c(1.6, 1.8)
toy$value[toy$promoter == "P1" & toy$srn_code == "C1"] <- c(4.5, 4.7)

toy$value[toy$promoter == "P2" & toy$srn_code == "DMSO1"] <- c(0.8, 1.0)
toy$value[toy$promoter == "P2" & toy$srn_code == "DMSO2"] <- c(0.9, 1.1)
toy$value[toy$promoter == "P2" & toy$srn_code == "C1"] <- c(1.5, 1.7)

evc <- fit_empty_vector_control(
  toy,
  response = "value",
  control = c("DMSO1", "DMSO2")
)

evc$replicate_results
evc$pair_results
```

## Local workflow equivalence check

Using the local Salmonella workflow files, `fit_empty_vector_control()` recovers
the original `04-hit_determination/hit_table.tsv.gz` output numerically:

```text
pair rows: 68238
max absolute delta in log.evcfc: 4.44e-16
max absolute delta in zscore:    3.55e-15
max absolute delta in pvalue:    1.22e-15
max absolute delta in BH pvalue: 2.11e-15
hit label mismatches: 0
```

Those values are floating-point roundoff, so this is effectively a one-to-one
reproduction of the original Salmonella workflow.
